---
title: "Open and Reproducible Science in R Module"
subtitle: "Induction Session"  
author: 
  - "Ernest Guevarra"
date: '10 October 2025'
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)

if (!require(remotes)) install.packages("remotes")
if (!require(fontawesome)) remotes::install_github("rstudio/fontawesome")
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_light(
  base_color = "#002147",
  title_slide_background_image = "",
  title_slide_background_size = "cover",
  header_font_google = google_font("Fira Sans"),
  text_font_google   = google_font("Fira Sans Condensed"),
  text_font_size = "1.2em",
  link_color = "#214700",
  header_h1_font_size = "50px",
  header_h2_font_size = "40px",
  header_h3_font_size = "30px",
  code_font_google   = google_font("Fira Mono"),
  text_slide_number_font_size = "0.5em",
  footnote_font_size = "0.5em"
)
```

# Outline

## Housekeeping

* What to expect from this module

* What is expected of you in this module

* Learning resources

## Practicals

* RStudio basics

* Some git and GitHub basics

---

class: inverse, center, middle

# What to expect from this module

---

# Module organisation

* The module is organised in two parts. 

* The first part focuses on teaching the basics of using R and RStudio. These topics are covered in the first 5 sessions of the module during Michaelmas term. 

* The second part, discussed in the last 5 sessions of the module during Hilary Term, focuses on giving on overview of open and reproducible science.

---

# Session organisation

* Each session is organised in two parts.

* The first part is theory - discussion on topic with slides.

* The second part is practical - coding exercises relevant to the topic 

---

# What is expected of you in this module

* Come to class on time.

* Participate in class particularly the exercises provided.

* Ask for help and be willing to help. 

---

# Learning resources

* Oxford iHealth Website - https://oxford-ihtm.io

* Oxford iHealth Handbook - https://oxford-ihtm.io/ihtm-handbook

* Module website - https://oxford-ihtm.io/open-reproducible-science

---

class: inverse, center, middle

# Thank you!

Slides can be viewed at https://oxford-ihtm.io/open-reproducible-science/session1.html

PDF version of slides can be downloaded at https://oxford-ihtm.io/open-reproducible-science/pdf/session1-getting-the-right-tools.pdf

R scripts for slides available [here](https://github.com/OxfordIHTM/open-reproducible-science/blob/main/session1.Rmd)
