---
title: "Data extraction, transformation, and loading in R"
subtitle: ""  
author: 
  - "Ernest Guevarra"
date: '2022-02-07'
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)

if (!require(remotes)) install.packages("remotes")
if (!require(fontawesome)) remotes::install_github("rstudio/fontawesome")
if (!require(tweetrmd)) remotes::install_github("gadenbuie/tweetrmd")
if (!require(webshot2)) remotes::install_github("rstudio/webshot2")
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_light(
  base_color = "#002147",
  title_slide_background_image = "",
  title_slide_background_size = "cover",
  header_font_google = google_font("Fira Sans"),
  text_font_google   = google_font("Fira Sans Condensed"),
  text_font_size = "1.2em",
  link_color = "#214700",
  header_h1_font_size = "50px",
  header_h2_font_size = "40px",
  header_h3_font_size = "30px",
  code_font_google   = google_font("Fira Mono"),
  text_slide_number_font_size = "0.5em",
  footnote_font_size = "0.5em"
)
```

# Outline

* Concepts and considerations about data

* Practical session

---

class: inverse, center, middle

# Concepts and considerations about data in tweets and quotes

---

class: inverse, center, middle

# Consideration #1:

# Data cleaning/processing/checking is an important, but very challenging, step when working with data

---

class: center, middle

<!---
# Data cleaning code cannot be clean. It's a sort of sin eater.
--->

```{r tweet8, echo = FALSE, eval = TRUE}
tweet_screenshot(
  tweet_url("StatFact", "492753200190341120"),
  hide_media = TRUE,
  theme = "dark"
)
```

[@StatFact](https://twitter.com/StatFact) [tweet](https://twitter.com/StatFact/status/492753200190341120)

---

class: center, middle

<!---
# I will let the data speak for itself when it cleans itself.
--->

```{r tweet1, echo = FALSE}
tweet_screenshot(
  tweet_url("AllisonReichel", "1350107719236247558"),
  #plain = TRUE,
  hide_media = TRUE,
  theme = "dark"
)
```

[@AllisonReichel](https://twitter.com/AllisonReichel)
[tweet](https://twitter.com/AllisonReichel/status/1350107719236247558)

---

class: inverse, center, middle

# and I’m still pretty sure some of the data is missing, but it could still be here, in this ONE HUNDRED SHEET excel file

```{r tweet6, echo = FALSE, eval = FALSE}
tweet_screenshot(
  tweet_url("RallidaeRule", "572922098617741312"),
  hide_media = TRUE,
  theme = "dark"
)
```

[@RallidaeRule](https://mobile.twitter.com/RallidaeRule) [tweet](https://mobile.twitter.com/rallidaerule/status/572922098617741312)

---

class: center, middle

<!----
# Never check data when you are hungry, thirsty, or tired.
--->

```{r tweet2, echo = FALSE}
tweet_screenshot(
  tweet_url("AmeliaMN", "1357861551575871494"),
  #plain = TRUE,
  hide_media = TRUE,
  theme = "dark"
)
```

[@GhazalGulati](https://twitter.com/GhazalGulati), via
[@AmeliaMN](https://twitter.com/AmeliaMN/)
[tweet](https://twitter.com/AmeliaMN/status/1357861551575871494)

---

class: inverse, center, middle

# Classroom data are like teddy bears; real data are like a grizzly with salmon blood dripping out its mouth.

```{r tweet5, echo = FALSE, eval = FALSE}
tweet_screenshot(
  tweet_url("sgrifter", "631150829165113344"),
  hide_media = TRUE,
  theme = "dark"
)
```

[@JennyBryan](https://twitter.com/jennybryan), via [@sgrifter](https://twitter.com/sgrifter) [tweet](https://twitter.com/sgrifter/status/631150829165113344)

---

class: center, middle

<!---
# Working with data is not about rules to follow but about decisions to make
--->

```{r tweet4, echo = FALSE}
tweet_screenshot(
  tweet_url("kwbroman", "686562071853547520"),
  hide_media = TRUE,
  theme = "dark"
)
```

[@naupakaz](https://twitter.com/naupakaz), via [@kwbroman](https://twitter.com/kwbroman) [tweet](https://twitter.com/kwbroman/status/686562071853547520)

---

class: inverse, center, middle

# Consideration #2

# Big data is what everyone is talking about

---

class: center, middle

<!---
# Big Data: (n): the belief that a big enough pile of horseshit will, with probability one, somewhere contain a pony.
--->

```{r tweet3, echo = FALSE}
tweet_screenshot(
  tweet_url("callin_bull", "831663695813963776"),
  hide_media = TRUE,
  theme = "dark"
)
```

[@mlipsitch](https://twitter.com/mlipsitch), via [@callin_bull](https://twitter.com/callin_bull) [tweet](https://twitter.com/callin_bull/status/831663695813963776)

---

class: inverse, center, middle

# Let's start the "titanic data" movement. Data too big to fail.

```{r tweet7, echo = FALSE, eval = FALSE}
tweet_screenshot(
  tweet_url("neilfws", "588533319135985665"),
  hide_media = TRUE,
  theme = "dark"
)
```

[@neilfws](https://twitter.com/neilfws) with assist from [@aaronquinlan](https://twitter.com/aaronquinlan), [tweet](https://twitter.com/neilfws/status/588533319135985665)

---

class: inverse, center, middle

# Consideration #3

# Data sharing requires a well-thought out process that everyone can follow

---

class: center, middle

<!---
# Last week I told a collaborator to stick the files on a USB drive and walk the 100m across the road rather than figure out inter-institute file sharing.
--->

```{r tweet9, echo = FALSE, eval = TRUE}
tweet_screenshot(
  tweet_url("PeteHaitch", "1095948744510517250"),
  hide_media = TRUE,
  hide_thread = TRUE,
  theme = "dark"
)
```

[@PeteHaitch](https://twitter.com/PeteHaitch) [tweet](https://twitter.com/PeteHaitch/status/1095948744510517250)

---

class: inverse, center, middle

# Consideration #4

# Working with data in R is great!

---

class: center, middle

<!---
# R is a datasmith's heaven-on-earth; I like Python, long term relationship with Excel, quite like Power Query, DAX's a keeper, but I love R.
--->

```{r tweet10, echo = FALSE, eval = TRUE}
tweet_screenshot(
  tweet_url("tggleeson", "591013073542062081"),
  hide_media = TRUE,
  theme = "dark"
)
```

[@tggleeson](https://twitter.com/tggleeson)
[tweet](https://twitter.com/tggleeson/status/591013073542062081)

---

class: inverse, center, middle

# Practical session

---

# Practical session

* We will all go through Exercise #1 in the [Practical R for Epidemiologists](https://practical-r.org/exercise1.html) book

* This is also one of your next assignment for this lecture series

---

class: inverse, center, middle

# Questions?

---

class: inverse, center, middle

# Thank you!

Slides can be viewed at https://OxfordIHTM.github.io/open-reproducible-science/session3.html

PDF version of slides can be downloaded at https://OxfordIHTM.github.io/open-reproducible-science/pdf/session3-working-with-data-in-r.pdf

R scripts for slides available [here](https://github.com/OxfordIHTM/open-reproducible-science/blob/82bfbf7cb5b71fcb4df9baf57f735482da78b498/session3.Rmd)

