---
title: "Creating your own functions"
subtitle: "Learning the basics of R - Part 3"  
author: 
  - "Ernest Guevarra"
date: '2023-11-27'
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)

if (!require(remotes)) install.packages("remotes")
if (!require(fontawesome)) remotes::install_github("rstudio/fontawesome")
if (!require(tweetrmd)) remotes::install_github("gadenbuie/tweetrmd")
if (!require(webshot2)) remotes::install_github("rstudio/webshot2")
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_light(
  base_color = "#002147",
  title_slide_background_image = "",
  title_slide_background_size = "cover",
  header_font_google = google_font("Fira Sans"),
  text_font_google   = google_font("Fira Sans Condensed"),
  text_font_size = "1.2em",
  link_color = "#214700",
  header_h1_font_size = "50px",
  header_h2_font_size = "40px",
  header_h3_font_size = "30px",
  code_font_google   = google_font("Fira Mono"),
  text_slide_number_font_size = "0.5em",
  footnote_font_size = "0.5em"
)
```

# Outline

* Why write functions

* When to write functions

* How to write functions

* Practical session

---

# Why write functions

* Allow automation of common tasks in a more powerful and general way than *copy-and-pasting*;

    * You can give a function an evocative name that makes your code easier to understand;
    * As requirements change, you only need to update code in one place, instead of many.
    * You eliminate the chance of making incidental mistakes when you copy and paste (i.e. updating a variable name in one place, but not in another).

* Fewer global variables: When you run a function, the intermediate variables that it creates are not stored in your global environment. This saves memory and keeps your global environment cleaner.

* Better documentation: Well documented functions help the user understand the steps of your processing.

* Easier to maintain / edit: When you create a function for a repeated task, it is easy to edit that one function. Then every location in your code where that same task is performed is automatically updated.

---

# When to write functions

.pull-left[
You should consider writing a function whenever you’ve copied and pasted a block of code more than twice (i.e. you now have three copies of the same code). 

For example, take a look at this code.

What does it do?
]

.pull-right[
```R
df <- data.frame(
  a = c(1, 2, 1, 1, 1, 2, 1, 1, 2, 2),
  b = c(2, 2, 2, 1, 1, 1, 1, 2, 1, 2),
  c = c(1, 2, 1, 1, 2, 1, 2, 1, 2, 2)
)

df$a <- ifelse(df$a == 2, 0, df$a)
df$b <- ifelse(df$b == 2, 0, df$b)
df$c <- ifelse(df$c == 2, 0, df$c)
```
]

---

# When to write functions

.pull-left[
Original data:

```{r, echo = FALSE}
df <- data.frame(
  a = c(1, 2, 1, 1, 1, 2, 1, 1, 2, 2),
  b = c(2, 2, 2, 1, 1, 1, 1, 2, 1, 2),
  c = c(1, 2, 1, 1, 2, 1, 2, 1, 2, 2)
)

df
```
]

.pull-right[
Recoded data:

```{r, echo = FALSE}
df$a <- ifelse(df$a == 2, 0, df$a)
df$b <- ifelse(df$b == 2, 0, df$b)
df$c <- ifelse(df$c == 2, 0, df$c)

df
```
]

--

This is a good example of when writing a function will be useful/beneficial.

---

# How to write functions

.pull-left[
We can create a function called `recode_values()`:

```R
recode_values <- function(x) { 
  ifelse(x == 2, 0, x)
}
```

And apply it to the same data as follows:

```R
df$a <- recode_values(df$a)
df$b <- recode_values(df$b)
df$c <- recode_values(df$c)
```
]

--

.pull-right[
We get:

```{r, echo = FALSE}
df <- data.frame(
  a = c(1, 2, 1, 1, 1, 2, 1, 1, 2, 2),
  b = c(2, 2, 2, 1, 1, 1, 1, 2, 1, 2),
  c = c(1, 2, 1, 1, 2, 1, 2, 1, 2, 2)
)

recode_values <- function(x) { 
  ifelse(x == 2, 0, x)
}

df$a <- recode_values(df$a)
df$b <- recode_values(df$b)
df$c <- recode_values(df$c)

df
```
]

---

# How to write functions

* You need to pick a name for the function. In the example I used `recode_values` because this function recodes the values based on a specified rule (i.e., value of 2 is converted to 0).

* You list the inputs, or **arguments**, to the function inside `function`. Here we have just one argument. If we had more the call would look like `function(x, y, z)`.

*  You place the code you have developed in body of the function, a `{` block that immediately follows `function(...)`.

???

Note the overall process: I only made the function after I’d figured out how to make it work with a simple input. It’s easier to start with working code and turn it into a function; it’s harder to create a function and then try to make it work.

---

class: inverse, center, middle

# Questions?

---

class: inverse, center, middle

# Practical session

We'll work through *Exercise 2 - Manipulating objects and creating new functions* in Practical R for Epidemiologists (https://practical-r.org/exercise2.html) as a GitHub Classroom assignment

---

class: inverse, center, middle

# Thank you!

Slides can be viewed at https://oxford-ihtm.io/open-reproducible-science/session4.html

PDF version of slides can be downloaded at https://oxford-ihtm.io/open-reproducible-science/pdf/session4-r-basics-part3.pdf

R scripts for slides available [here](https://github.com/OxfordIHTM/open-reproducible-science/blob/main/session4.Rmd)

